Neural fields with different embedding dimensionality can express identical dynamics

### 2024-02-23 paper/manuscript.tex
henlo yes i did the coarse-graining stuffs

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}
Classical neural field models describe interactions between neurons positioned in an embedding; often a physical space (for instance, that of the cortical sheet) or, later, abstract spaces (for instance, that of neuron celltypes), in which the dynamics are highly regular in space and time. Recent improvements in recordings of neural activity have further demonstrated the smoothness of the neural dynamics, when placed in an appropriate embedding space.

@@

[TODO: add the pretty figure here]

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
[TODO]
\linenumbers

@@

\subsection{Motivation for a grid-based approach}

@@

Sampling the original connectivity kernel at the grid points $\vec{v_i}$ and $\vec{v_j}$ yields a numerical connectivity matrix $J_{ij} := w_U(\vec{v_i}, \vec{v_j})$, illustrated in Fig. [TODO: add a figure here of the discretization here and the sampling on a grid, and the limit of n to infty. do we need to clarify explicitly the bounds of each square bin ?], which can be used to simulate the resulting population dynamics equation:

@@

@@

Let us define a sequence of bijective mappings $\{S^n\}_{n \in \mathbb N}$, acting on the $n$-bit truncations $\vec{v^{(n)}}$ of any $\vec{v} \in [0,1]^2$, as well as the corresponding sequence of images $\{\alpha^n := S^n(\vec{v^{(n)}})\}_{n \in \mathbb N}$. $S^n$ defines the following bijection between points on the two-dimensional grid and the one-dimensional line:

\begin{equation} \label{eq:def-S} \begin{aligned}
S^n :& \{\tfrac{j-1}{2^n}\;|\;j\in\{1,\cdots,2^n\}\}^2 \mapsto \{\tfrac{i-1}{4^n}\;|\;i\in\{1,\cdots,4^n\}\},\\
\text{such that }&(v_1^{(n)}, v_2^{(n)}) = (0.b_1^1 b_1^2 \cdots b_1^n, 0.b_2^1 b_2^2 \cdots b_2^n) \mapsto \alpha^n = 0.b^1 b^2 \cdots b^{2n}.
\end{aligned} \end{equation}

@@

However, for any $(v_1,v_2)$, the sequence $\{A^n(v_1, v_2)\}_{n \in \mathbb N}$ does not converge. In Fig. [TODO: fig 3.8 of the thesis] we graph the mapping $A^n$, which demonstrates that the image of any point $(v_1,v_2)$ in $[0,1]^2$ diverges as $n$ increases. As $n$ increases, $A^n$ ``refines'' its input with fluctuations of order $2^{-n}$ (because it works with truncations of $(v_1, v_2)$, which have error $c^{(n)} < 2^{-n}$), but the image $\alpha^n=b_1^n 2^{-1} + b_2^n 2^{-2} + b_1^{n-1} 2^{-3} + b_2^{n-1} 2^{-4} \cdots$ fluctuates (at least) with fixed amplitude $\sim 2^{-2}$, independantly of $n$. Therefore the image never converges, and $\{A_n\}_{n \in \mathbb N}$ is not pointwise convergent.

@@

@@

@@

@@

introduce coarse-graining procedure (applying the smoothness hypothesis numerically)

@@

% TODO: are all bijective limits local ? does local imply bijection ?
% in this case, we have shown that bijection + locality => we can safely do the mapping (bijection for the analytical approach, locality for the numerical approach)

>>>>

\ifndef{SUBMISSION}\usepackage{graphicx}\endif

@@

\end{flushleft}
% Please keep the abstract below 300 words
\section*{Abstract}

\ifndef{SUBMISSION}
\includegraphics[width=\textwidth]{../figures/cover.pdf}
\endif

Classical neural field models describe interactions between neurons positioned in an embedding; often a physical space (for instance, that of the cortical sheet) or, later, abstract spaces (for instance, that of neuron celltypes), in which the dynamics are highly regular in space and time. Recent improvements in recordings of neural activity have further demonstrated the smoothness of the neural dynamics, when placed in an appropriate embedding space.

@@

% Please keep the Author Summary between 150 and 200 words
% Use first person. PLOS ONE authors please skip this step. 
% Author Summary not valid for PLOS ONE submissions.   
\section*{Author summary}
[TODO?]
\linenumbers

@@

\subsection{Motivation for a grid-based approach} \label{sec:grid-motivation}

@@

Sampling the original connectivity kernel at the grid points $\vec{v_i}$ and $\vec{u_j}$ yields a numerical connectivity matrix $J_{ij} := w_U(\vec{v_i}, \vec{u_j})$, illustrated in Fig. [TODO], which can be used to simulate the resulting population dynamics equation:

@@

\ifndef{SUBMISSION}
\includegraphics[height=20em]{../figures/discretization.png}

[TODO: the discretization here and the sampling on a grid, and the limit of n to infty. do we need to clarify explicitly the bounds of each square bin ?]
\endif

@@

Let us define a sequence of bijective mappings $\{S^n\}_{n \in \mathbb N}$, acting on the $n$-bit truncations $\vec{v^{(n)}}$ of any $\vec{v} \in [0,1]^2$, as well as the corresponding sequence of images $\{\alpha^n := S^n(\vec{v^{(n)}})\}_{n \in \mathbb N}$. $S^n$ defines the following bijection between points on the two-dimensional and one-dimensional grids:

\begin{equation} \label{eq:def-S} \begin{aligned}
S^n :& \{\tfrac{l-1}{2^n}\;|\;l\in\{1,\cdots,2^n\}\}^2 \mapsto \{\tfrac{i-1}{4^n}\;|\;i\in\{1,\cdots,4^n\}\},\\
\text{such that }&(v_1^{(n)}, v_2^{(n)}) = (0.b_1^1 b_1^2 \cdots b_1^n, 0.b_2^1 b_2^2 \cdots b_2^n) \mapsto \alpha^n = 0.b^1 b^2 \cdots b^{2n}.
\end{aligned} \end{equation}

@@

However, for any $(v_1,v_2)$, the sequence $\{A^n(v_1, v_2)\}_{n \in \mathbb N}$ does not converge. In Fig. [TODO] we graph the mapping $A^n$, which demonstrates that the image of any point $(v_1,v_2)$ in $[0,1]^2$ diverges as $n$ increases. As $n$ increases, $A^n$ ``refines'' its input with fluctuations of order $2^{-n}$ (because it works with truncations of $(v_1, v_2)$, which have error $c^{(n)} < 2^{-n}$), but the image $\alpha^n=b_1^n 2^{-1} + b_2^n 2^{-2} + b_1^{n-1} 2^{-3} + b_2^{n-1} 2^{-4} \cdots$ fluctuates (at least) with fixed amplitude $\sim 2^{-2}$, independantly of $n$. Therefore the image never converges, and $\{A_n\}_{n \in \mathbb N}$ is not pointwise convergent.

\ifndef{SUBMISSION}
\includegraphics[width=\linewidth]{../figures/mapping-A.pdf}
[TODO: remove the numbers on this figure, dot the grid nodes]
\endif

@@

\ifndef{SUBMISSION}
\includegraphics[height=20em]{../figures/mapping-col-Z.pdf}

[TODO: add labels, show n=3, etc.]
\endif

@@

In the interest of developing an insight on the properties of the two mappings introduced previously, let us now develop a numerical grid-based simulation.
As established in the \autoref{sec:grid-motivation}, the original two-dimensional neural field can be simulated by sampling the connectivity kernel at pairs of discrete grid points $(\vec{v_i},\vec{u_j})$, discretizing the neural field to a set of $4^n$ scalars $\{h_i(t) \;|\; i=1,\cdots,4^n\}$.

Naively, we could proceed in the same way to simulate the corresponding one-dimensional neural field, by sampling the connectivity kernel $\tilde w(\alpha_i, \beta_j) = w_U(S^{-1}(\alpha_i), S^{-1}(\beta_j)) = w_U(\vec{v_i}, \vec{u_j})$, where the pairs of discrete one-dimensional points $(\alpha_i,\beta_j)$ are obtained by mapping the two-dimensional grid-points $(\vec{v_i},\vec{u_j})$.

Critically, since $S$ defines a bijection, it acts as a permutation on the discrete connectivity matrix indices. Furthermore the population dynamics in \autoref{eq:population-dynamics} are invariant to permutation in the indices, and the naively sampled connectivity matrix corresponding to the one-dimensional neural field trivially yields identical dynamics.

In essence, since both connectivity matrices of the one- and two-dimensional neural fields have the same size, we are compensating for the reduction in dimensionality by increasing the one-dimensional grid resolution. In two dimensions, the $2^n \times 2^n = 4^n$ grid points are spaced with $2^{-n}$ along every axis, but after the mapping, those $4^n$ grid points are spaced every $4^{-n}$, since there are twice as many bits to describe the position along the unique axis. This is illustrated in Fig. [TODO].

\ifndef{SUBMISSION}
\includegraphics[width=\linewidth]{../figures/S-viz.export.pdf}
[TODO: adapt this figure better, but gives the idea]
\endif

% We then loosely have the following correspondence between the discrete connectivity matrices for the one- and two-dimensional neural fields:

% \begin{equation} \begin{aligned}
% \tilde w(\alpha_i,\beta_j) &= \tilde J_{\alpha_i,\beta_j} \\
% &= \tilde J_{S(\vec{v_i},\vec{u_j})} \\
% &= J_{ij} \;\text{by definition} \\
% &= w_U(\vec{v_i},\vec{u_j})
% \end{aligned}\end{equation}

@@

Coarse-graining adresses the problem of numerically enforcing the mean-field approximation. The mean-field approximation is done on the original two-dimensional neural field when sampling the connectivity matrix $J_{ij}$, as all neurons inside the same grid cells are assumed to be the same. This is not the case when defining the connectivity of the one-dimensional neural field as the permutation of rows and columns of $J_{ij}$.

In order to perform the mean-field approximation on the one-dimensional neural field, we average together $2^n$ consecutive one-dimensional bins, each of length $4^{-n}$, resulting in $2^n$ coarse-grained bins. This is illustrated in Fig. [TODO] Specifically, for every coarse-grained bin $a=1,\cdots,2^n$, the set of consecutive positions $\mathcal{C}_a = \{\tfrac{a-1}{2^n} + \tfrac{l-1}{4^n} \;|\; l \in 1, \cdots, 2^n\}$, gets mapped to the two-dimensional grid points and averaged:

\begin{equation*}
\vec{\bar v_a} := \textrm{mean}(\{S^{-1}(\alpha_l) \;|\; \alpha_l \in \mathcal{C}_a\}).
\end{equation*}

\ifndef{SUBMISSION}
\includegraphics[height=15em]{../figures/coarse-graining.pdf}

[TODO: add the indices a,b and l on this figure, for clarity.]
\endif

The obtained coarse-grained points are then used to sample the connectivity kernel, and form the connectivity matrix of the one-dimensional neural field:

\begin{equation}
\tilde J_{ab} := w_U(\vec{\bar v_a}, \vec{\bar v_b}).
\end{equation}

@@

@@

% TODO: are all bijective limits local ? does local imply bijection ?
% in this case, we have shown that bijection + locality => we can safely do the mapping (bijection for the analytical approach, locality for the numerical approach)
% TODO: mention the "standard example" of the Z-mapping
